Solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit

### 2016-08-09 paper/paper.tex
Misc edits

\begin{document}
\title{Average solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit}
\author{V. Veibell and R.S. Weigel}
\doublespacing
\maketitle

@@

%\twocolumn[
%  \begin{@twocolumnfalse}
%\maketitle
%\hrule
\begin{abstract}
We consider two types of events, identified by decreases in $D_{st}$ below a threshold value and increases in the the equatorial mass density at geosynchronous altitudes, $\rho_{eq}$, above a threshold value on 1-hour and 1-day time scales using the \cite{Takahashi2010} dataset.  From the $D_{st}$ events, we find that there is a statistically weak and small amplitude difference between $\rho_{eq}$ on the day of the event and the days before and after on daily timescales.  On hourly timescales, it is observed that for 6 hours after onset, very few measurements of  $\rho_{eq}$ are found and that most of the contribution to the daily averages is from 6-24 hours after onset.  On hourly timescales, $\rho_{eq}$ following the onset of a $D_{st}$ event does not depend on the north-south component of the interplanetary magnetic field, $B_z$ near onset time. The primary factor that determines if a post-$D_{st}$ event onset peak in $\rho_{eq}$ will occur is elevated $F10.7$.  From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ four hours after the event onset corresponding to larger $\rho_{eq}$ values 24-36 hours after onset.  In contrast, the average of $B_z$ four hours prior to $\rho_{eq}$ events does not have a statistically significant impact on $\rho_{eq}$ after onset.
\end{abstract}
%\vspace{2em}
%\hrule
%\vspace{2em}
% \end{@twocolumnfalse}
%  ]

@@

Our first analysis uses GOES-6 data only from the time interval 1989-1991, which corresponds with the interval used in the epoch analysis of \cite{Takahashi2010}. In this interval, there were 75 $D_{st}$ events where $D_{st}$ remained below the $-50$~nT threshold for at least 12 hours. Only events with an onset time between 06 and noon MLT were considered. For both types of events, the first hour of a local minimum in $D_{st}$ following a threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after event onset. For multiple detected threshold crossings within 24 hours of each other, the first crossing was selected as the event since visual inspection showed the following crossings to typically be from brief deviations during the recovery period of a geomagnetic storm. To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across events on each epoch day.

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for the 75 events from 1989-1991.  The minimum $D_{st}$ median is $-75$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$ before and after the event, although the magnitude of increase observed here is, on average, $4.5$ amu/cm$^3$ (from $\sim$19 to ${\sim}$24~amu/cm$^3$) instead of the increase of ${\sim} 10$ amu/cm$^3$ found in Figure 11 of \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute the medians and red error lines.  Error lines for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values. For the non-$\rho_{eq}$\ parameters, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of $\rho_{eq}$\ values as all available measurements were used instead of restricting to only values where a $\rho_{eq}$ value also existed.

@@

If a statistically significant difference in the medians between two days existed, the fraction of bootstrap samples with a different sign is expected to be much smaller ($<$ 5\%).  In terms of a hypothesis test, we cannot reject the null hypothesis that the medians are the same with a confidence higher than 100 minus the \% value shown in the table; the highest confidence level of 86\% is in the difference between day 0 and day -1.

@@

These tests indicate that there is not a statistically significant difference in medians of daily averaged $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.  In addition, the observed differences are on the order of only 20\% of the value of $\rho_{eq}$ on the onset day, which is similar to the variation in the observations over the displayed epoch time interval.

@@

We have also considered the hourly averages from the events shown in the bottom panel of Figure~\ref{DailyAveragedDstEvents}, which shown in the top panel of Figure~\ref{HourlyAveragedDstEvents}.  In the time period of elevated mass density within 6 hours of the start of the event, there are only $\sim$4 samples per hourly interval.  The peak value of $\sim$70 amu/cm$^3$ is associated with a single observation.

To increase the number of events, we have removed the constraint that the events must occur in the MLT range from 06-12 and the constraint that $D_{st}$ remained below the threshold value for 12 hours or longer.  The result is shown in the bottom panel of Figure~\ref{HourlyAveragedDstEvents}. The two spikes (one updwards after onset, one downwards before onset) were tested for significant deviation from two baselines, one 24-12 hours before onset, one 16-48 hours after onset. The probability of the left spike coming from the same distribution as the left baseline is 1.06\%, and 8.2\% from the right baseline, whereas the probability of the right spike coming from the left baseline is 1.89\% and 0.17\% from the right baseline.

%This result is consistent with previous observations of large mass density increases in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}).

@@

To determine if there is an IMF $B_{z}$ dependence, the epoch curves in the top panel of Figure~\ref{HourlyAveragedDstEvents} for $\rho_{eq}$ and $B_{z}$ were separated into two parts depending whether $B_z$ was above or below its average four hours before and including the onset hour and four hours after and including the hour of onset; the result is shown in Figure~\ref{fig:BzBinnedDstEvents}. Any statistically significant difference at a 95\% confidence level between the two parts is indicated by a green dot on the horizontal axis using the same bootstrap method described previously.  (At this confidence level, false positives are expected to randomly occur 5\% of the time, or 3.6 times for 72 values.)  These two curves indicate that there is not a statistically significant dependence on IMF $B_{z}$ around the time of the onset of a $D_{st}$ event.

The process for creating Figure~\ref{fig:BzBinnedDstEvents} was repeated except using the sort variable $F10.7$ instead of $B_z$, as shown in Figure~\ref{fig:F107BinnedDstEvents}. As expected, difference between the epoch curves is statistically significant at all times, as expected, but elevated $F10.7$ is associated with a peak in the $\rho_{eq}$ curve approximately 5 hours after onset. The values contributing to the peak were compared to a baseline of all values 16-48 hours after onset via a Wilcoxon rank-sum test and found to have significantly different medians at the 99\% significance level.

@@

An alternative type of event can be defined that corresponds to large increases in $\rho_{eq}$.  Figure~\ref{HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\rho_{eq}$ above $20$~amu/cm$^2$ defines the start of the event. In finding $\rho_{eq}$ events, missing values were replaced with linearly interpolated values.  These events are associated with positive $B_z$ and low values of $D_{st}$.  Approximately 8 hours after onset, $\rho_{eq}$ peaks and then slowly decays after 40 hours to pre--onset levels.

@@

The largest difference occurs when the separation is performed after onset, with more positive $B_z$ leading to larger peak $\rho_{eq}$ than more negative $B_{z}$.  The rate of change of $\rho_{eq}$ also appears to be insenstive to $B_z$ for approximately 12 hours prior to onset.  The equivalent epoch separation curves for $P_{dyn}$ had a statistically significant differences similar to that for $B_z$, with higher dynamic pressure playing a role similar to positive $B_z$.

@@

On average, around the time of large geomagnetic events, there is a very weak enhancement of $\rho_{eq}$ and this enhancement does not depend on either the interplanetary magnetic field $B_z$ or the intensity of the geomagnetic event.  This indicates that there is an approximate balance between magnetospheric enhancement in $\rho_{eq}$ and any reduction due to the inward motion of the plasmasphere.

$\rho_{eq}$ events are larger when $B_{z}$ is more positive after the onset of the event, which indicates that a substantial portion of the events may due to the apparent mass refilling mechanism described by \cite{Denton2016}.  Without additional information about the location of the plasmapause, it may not be possible to separate out the conditions required for enhancement due to enhanced magnetospheric activity from this mechanism.

>>>>

\begin{document}
\title{Solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit}
\author{V. Veibell and R.S. Weigel}
\doublespacing
\maketitle

@@

%\twocolumn[
%  \begin{@twocolumnfalse}
%\maketitle
%\hrule
\begin{abstract}
We consider two types of events, identified by decreases in $D_{st}$ below a threshold value and increases in the the equatorial mass density at geosynchronous altitudes, $\rho_{eq}$, above a threshold value using the \cite{Takahashi2010} dataset.  From the $D_{st}$ events and 1-day averages, we find that there is a statistically weak and small amplitude difference between $\rho_{eq}$ on the day of the event and the days before and after.  For hourly averages, it is found that for 6 hours after onset, very few measurements of $\rho_{eq}$ exist and that most of the contribution to the daily averages is from 6-24 hours after onset.  For hourly averages, $\rho_{eq}$ following the onset of a $D_{st}$ event does not depend on the north-south component of the interplanetary magnetic field, $B_z$ near onset time. The primary factor that determines if a post-$D_{st}$ event onset peak in $\rho_{eq}$ will occur in the hourly averages is elevated $F10.7$.  From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ four hours after the event onset corresponding to larger $\rho_{eq}$ values 24-36 hours after onset.  In contrast, the average of $B_z$ four hours prior to $\rho_{eq}$ events does not have a statistically significant impact on $\rho_{eq}$ after onset.
\end{abstract}
%\vspace{2em}
%\hrule
%\vspace{2em}
% \end{@twocolumnfalse}
%  ]

@@

Our first analysis uses GOES-6 data only from the time interval 1989-1991, which corresponds with the interval used in the epoch analysis of \cite{Takahashi2010}. In this interval, there were 75 $D_{st}$ events where $D_{st}$ remained below the $-50$~nT threshold for at least 12 hours. Only events with an onset time between 06:00 and 12:00 MLT were considered. The first hour of a local minimum in $D_{st}$ following a threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after event onset. For multiple detected threshold crossings within 24 hours of each other, the first crossing was selected as the event because visual inspection showed the following crossings to typically be from brief downward deviations during the recovery period of a geomagnetic storm. To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across events on each epoch day.

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for the 75 events from 1989-1991.  The minimum $D_{st}$ median is $-75$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$ before and after the event, although the magnitude of increase observed here is $4.5$ amu/cm$^3$ instead of the increase of ${\sim} 10$ amu/cm$^3$ found in Figure 11 of \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute the medians and red error lines.  Error lines for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values. For the non-$\rho_{eq}$\ parameters, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of $\rho_{eq}$\ values as all available measurements were used instead of restricting to only values where a $\rho_{eq}$ value also existed.

@@

If a statistically significant difference in the medians between two days existed, the fraction of bootstrap samples with a different sign is expected to be much smaller ($<$ 5\%).  In terms of a hypothesis test, we cannot reject the null hypothesis that the medians are the same with a confidence higher than 100 minus the percentage value shown in the table; the highest confidence level of 86\% is in the difference between day 0 and day -1.

@@

These tests indicate that there is not a statistically significant difference in medians of daily-averaged $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.  In addition, the observed differences are on the order of only 20\% of the value of $\rho_{eq}$ on the onset day, which is similar to the variation in the observations over the displayed epoch time interval.

@@

We have also considered the hourly averages from the events shown in the bottom panel of Figure~\ref{DailyAveragedDstEvents}, which shown in the top panel of Figure~\ref{HourlyAveragedDstEvents}.  In the time period of elevated mass density within 6 hours of the start of the event, there are only $\sim$4 samples per hourly interval and the peak value of $\sim$70 amu/cm$^3$ is associated with a single observation.

To increase the number of events, we have removed the constraint that the events must occur in the MLT range from 06:00-12:00 and the constraint that $D_{st}$ remained below the threshold value for 12 hours or longer.  The result is shown in the bottom panel of Figure~\ref{HourlyAveragedDstEvents}. The two spikes (one upwards after onset, one downwards before onset) were tested for significant deviation from two baselines, one 24-12 hours before onset, one 16-48 hours after onset. The probability of the left spike coming from the same distribution as the left baseline is 1.06\%, and 8.2\% from the right baseline, whereas the probability of the right spike coming from the left baseline is 1.89\% and 0.17\% from the right baseline.  This indicates that there is a statistically significant peak in $\rho_{eq}$ within 6 hours of onset when hourly averages are considered.  This result is consistent with previous observations of large mass density increases in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}).

@@

To determine if there is an IMF $B_{z}$ dependence, the epoch curves in the bottom panel of Figure~\ref{HourlyAveragedDstEvents} for $\rho_{eq}$ and $B_{z}$ were separated into two parts depending whether $B_z$ was above or below its average four hours before and including the onset hour and four hours after and including the hour of onset; the result is shown in Figure~\ref{fig:BzBinnedDstEvents}. Any statistically significant difference at a 95\% confidence level between the two parts is indicated by a green dot on the horizontal axis using the same bootstrap method described previously.  (At this confidence level, false positives are expected to randomly occur 5\% of the time, or 3.6 times for 72 values.)  These two plots indicate that there is not a statistically significant dependence on IMF $B_{z}$ around the time of the onset of a $D_{st}$ event.

The process for creating Figure~\ref{fig:BzBinnedDstEvents} was repeated except using the sort variable $F10.7$ instead of $B_z$, and the result is shown in Figure~\ref{fig:F107BinnedDstEvents}. The difference between the epoch curves is statistically significant at all times, as expected due to the high correlation between $\rho_{eq}$ and $F10.7$, but elevated $F10.7$ is associated with a peak in the $\rho_{eq}$ curve approximately 5 hours after onset. The values contributing to the peak were compared to a baseline of all values 16-48 hours after onset via a Wilcoxon rank-sum test and were found to have significantly different medians at the 99\% significance level.  This indicates the possibility that there must be a high base-line level of $\rho_{eq}$ in order for an increase in $\rho_{eq}$ to be observed for a $D_{st}$ event.

@@

An alternative type of event can be defined that corresponds to large increases in $\rho_{eq}$.  Figure~\ref{HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\rho_{eq}$ above $20$~amu/cm$^2$ defines the start of the event. In finding $\rho_{eq}$ events, missing values were replaced with linearly interpolated values.  These events are associated with positive $B_z$ and low values of $D_{st}$.  Approximately 12 hours after onset, $\rho_{eq}$ slowly decays over 40 hours to near pre--onset levels.

@@

Figure~\ref{BzBinned} shows the the events of Figure~\ref{HourlyAveragedRhoEvents} when separated by $B_z$ in the same manner as Figure~\ref{fig:BzBinnedDstEvents}.  The largest difference occurs when the separation is performed after onset, with more positive $B_z$ leading to a larger peak in $\rho_{eq}$ than more negative $B_{z}$.  The rate of change of $\rho_{eq}$ also appears to be insenstive to $B_z$ for approximately 12 hours prior to onset.

%The equivalent epoch separation curves for $P_{dyn}$ had a statistically significant differences similar to that for $B_z$, with higher dynamic pressure playing a role similar to positive $B_z$.

@@

On average, around the time of large geomagnetic events, there is a statistically weak enhancement of daily-averaged $\rho_{eq}$, but consideration of hourly average $\rho_{eq}$ shows enhancements over a 6-hour window that contains few events.  For the hourly-averaged measurements, the average of $B_{z}$ around onset time did not have a significant impact on $\rho_{eq}$ after onset.

It was shown that a key factor in determining if large $\rho_{eq}$ enhancements will be observed is elevated $F10.7$.  When geomagnetic events were separated according to if they occured when $F10.7$ was low or high, significant enhancements in $\rho_{eq}$ were only observed under high $F10.7$ conditions.

For $\rho_{eq}$, more postive $B_{z}$ after onset lead to higher values of subsequent $\rho_{eq}$, which indicates that a substantial portion of the events may due to the apparent mass refilling mechanism described by \cite{Denton2016}.  Without additional information about the location of the plasmapause, it may not be possible to separate out the conditions required for enhancement due to enhanced magnetospheric activity from this mechanism.
